Exploiting Intrinsic Noise to Achieve Heterogeneous Control of Cell Differentiation Dynamics in the Presence of Time Delays, Model Uncertainties, and Extrinsic Variability

### 2023-07-11 paper.tex
Small edits.

We consider two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that bases its control signal on simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies only on observations from a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2),
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_2),
\end{equation}
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two.

@@

In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection\cite{XXX}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}. 
Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration. 
To explore how mismatches in assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model. 
To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
\begin{equation}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{eqnarray}
w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
 w'_2(\mathbf{X}) &= \gamma x_1/\alpha.
\end{eqnarray}
We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
%\begin{equation}
%u'=u(x_0 / \alpha, x_1 / \alpha) 
%\end{equation}
%\begin{equation}
%u'=u(\alpha x_0)
%\end{equation}
Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes. For example, in the two cell system the new steady state performance error would become:
\begin{eqnarray}
(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*.
\label{EuclidV}
\end{eqnarray}
We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and these are not recomputed or refit upon changing the system granularity.

@@

\subsection{Effects of changes to system granularity}
\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{Granularity.pdf}
\vspace{-0.1in}
\caption{Systems with increased granularity are less noisy and have better control performance. (A-C) Joint (left) and marginal (right) distributions for the FAC shows increased control performance and tighter distributions as $\alpha$ increases from 0.2 (top row) to 2.0 (bottom row) . (D-F) Joint (left) and marginal (left) distributions using the PAC controller at different levels of granularity.}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}
% XXX - This figure needs larger fonts for all the labels and axes. 
% MMM - made the change
% Update panel letters as written in caption
% Add tic marks to left side of the A and C heat maps
Stochastic fluctuations in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}. As the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate attractors for different steady state behaviors \cite{XXX}. This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved. The impact of reducing noise through increased system granularity on control performance remains unclear.

To gain insight on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0. 
% XXX - Why only go up to 2?  Did it become intractable?
% MMM - Non_Sparse Calculations are intractible after 2, but sparse versions of code will work fine
Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at low granularity ($\alpha=0.2$), original granularity ($\alpha=1.0$), and increased granularity ($\alpha=2.0$). As the system increases, the scale the size of the dynamics of the system becomes streched over a larger region in state space. This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller.

The noise reducing effect of granularity also improves the control performance of both controllers. Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller. This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate. 
% XXX - I would think you should be able to compute larger values of 10^6.  They would be slow, but should be doable in a few minutes per calculation.
% MMM - constructing the FSP takes longer than solving it
To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and the calculate the performance score. The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G. Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.

>>>>

We consider two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that bases its control signal on simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies only on observations from a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2),
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_2),
\end{equation}
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. Despite their differences in their observation data, both the FAC and the simpler PAC are optimized to achieve the same goal, namely to drive both cells to their respective set points.

@@

In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection\cite{XXX}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}. 
Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration. 
To explore how mismatches in assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model. 
To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
\begin{equation}\label{eq:granscale}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{eqnarray}
w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
 w'_2(\mathbf{X}) &= \gamma x_1/\alpha.
\end{eqnarray}
We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
%\begin{equation}
%u'=u(x_0 / \alpha, x_1 / \alpha) 
%\end{equation}
%\begin{equation}
%u'=u(\alpha x_0)
%\end{equation}
Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes. For example, in the two cell system the new steady state performance error would become:
\begin{eqnarray}
(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*.
\label{EuclidV}
\end{eqnarray}
We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and these are not recomputed or refit upon changing the system granularity.

@@

\subsection{Controllers trained using one assumed level of granularity can remain effective at other levels of granularity.}

We next explored how the granularity of the system would affect the differential control of the process. Specifically, we define the level of granularity using the parameter $\alpha$ as described in Section \ref{sec:Scaling} and which affects the propensity functions and scoring according to equations \ref{eq:granscale}-\ref{EuclidV}. Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to smaller population sizes (e.g., smaller volumes). We optimized the controller based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.

To explore this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.  Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at a low granularity ($\alpha=0.2$), the original granularity ($\alpha=1.0$), and at an increased granularity ($\alpha=2.0$).

As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
\brian[the previous two sentences do not make sense to me.]

The effect that granularity has on the control performance depends on two competing phenomena: First, because the controller was optimized for one granularity ($\alpha = 1$), one would expect that the controller would become worse if that granularity were in correct. However, at large granularity, we find that the opposite is true - the controller actually improves when applied to an incorrect model. The reason for this potentially surprising result, is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decrease with the inverse square root of the process scale \cite{Kampen1961}. At an extreme, as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that separate different steady state behaviors \cite{XXX}. This creates a tradeoff in noise-induced control, where noise enables differential control but is also deleterious to maintaining desired states once they have been achieved.

\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{Granularity.pdf}
\vspace{-0.1in}
\caption{Systems with increased granularity are less noisy and have better control performance. (A-C) Joint (left) and marginal (right) distributions for the FAC shows increased control performance and tighter distributions as $\alpha$ increases from 0.2 (top row) to 2.0 (bottom row) . (D-F) Joint (left) and marginal (left) distributions using the PAC controller at different levels of granularity.}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}

The noise reducing effect of granularity also improves the control performance of both controllers. Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller. This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate. 
To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and the calculate the performance score. The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G. Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.
